---
title: "Bioinf 524 - Final Project"
author: "Sumeed Manzoor"
date: today

format:
  pdf:
    pdf-engine: xelatex
    documentclass: article
    classoption: [11pt]
    geometry: margin=1in
    colorlinks: true
    linkcolor: maize
    urlcolor: maize
    citecolor: maize
    toc: true
    toc-depth: 1
    number-sections: true

mainfont: "TeX Gyre Pagella"
sansfont: "TeX Gyre Heros"
monofont: "Fira Code"

header-includes:
  - \usepackage{xcolor}
  - \definecolor{michiganblue}{HTML}{00274C}
  - \definecolor{maize}{HTML}{FFCB05}

  # Section title styling
  - \usepackage{sectsty}
  - \sectionfont{\color{michiganblue}}
  - \subsectionfont{\color{michiganblue}}
  - \subsubsectionfont{\color{michiganblue}}

  # Title styling
  - \usepackage{titling}
  - \pretitle{\begin{center}\color{michiganblue}\LARGE\bfseries}
  - \posttitle{\par\end{center}\vskip 0.5em}
  - \preauthor{\begin{center}\color{black}\large}
  - \postauthor{\par\end{center}}
  - \predate{\begin{center}\color{black}\small}
  - \postdate{\par\end{center}}

  # Bold text color override
  - \let\oldtextbf\textbf
  - \renewcommand{\textbf}[1]{\oldtextbf{\textcolor{michiganblue}{#1}}}

  # Hyperref setup (links)
  - \usepackage{hyperref}
  - \hypersetup{
      colorlinks=true,
      linkcolor=maize,
      urlcolor=maize,
      citecolor=maize
    }

  # Optional: nicer spacing
  - \usepackage{setspace}
  - \setstretch{1.1}

  # Optional: paragraph spacing instead of indent
  - \setlength{\parindent}{0pt}
  - \setlength{\parskip}{6pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

First, sorry for the spaghetti code. I tried to refactor and clean up the document, but for the most part it's quite raw.

# Libraries {.unnumbered}

```{r libraries}
# I learned that vscode will not respect this no matter what if I am using radian
# So on vscode, use base R.
Sys.setenv(
  RETICULATE_PYTHON = "/home/yoyomanzoor/.local/share/mamba/envs/py3.13/bin/python"
)
library(reticulate)
use_python(
  "/home/yoyomanzoor/.local/share/mamba/envs/py3.13/bin/python",
  required = TRUE
)

library(ggplot2)
library(ggpubr)
library(ggsci)
library(dplyr)
library(magrittr)
library(tidyr)
library(readr)
library(stringr)
library(knitr)
library(kableExtra)
library(patchwork)
library(tibble)
library(pheatmap)

library(zellkonverter)
library(SingleCellExperiment)
library(scater)
library(scran)
library(scuttle)
library(edgeR)
library(fgsea)
library(msigdbr)

theme_set(theme_pubr())
options(
  ggplot2.discrete.colour = scale_color_nejm,
  ggplot2.discrete.fill   = scale_fill_nejm
)

nejm_red  <- pal_nejm()(2)[1]  # "#BC3C29FF"
nejm_blue <- pal_nejm()(2)[2]  # "#0072B5FF"
```

# Set up

## Downloading data

The dataset for the whole brain CRISPR atlas is hosted on [huggingface](https://huggingface.co/datasets/perturbai/wholebrain_crispr_atlas). It's a large dataset -- 80 gigs!

To download, you should clone their [repo](https://github.com/jinlabneurogenomics/wholebrainperturbseq).

When I tried to use their scripts, however, I found some bugs. Since there is no license yet for me to copy their files with my edits, instead I will describe the changes you might need to make to their scripts and how to run them. Good luck!

At a high level, the way to use the scripts is run

```bash
python scripts/download.py --output-dir /path/to/output
```

You will need huggingface_hub installed from pip. It will download the h5ad shards, metadata, etc.

Then, you can run the preprocessing scripts they provided, with the following changes. In `merge_batches.py`, change the following:

```diff
    parser.add_argument(
        "--write_zarr",
-        type=bool,
-        default=False,
+        action="store_true",
        help="Write the merged adata to a zarr file",
    )
    parser.add_argument(
        "--write_h5ad",
-        type=bool,
-        default=False,
+        action="store_true",
        help="Write the merged adata to a h5ad file",
    )
```

In `processing_script.sh`, change the following:

```diff
diff --git a/wholebrainperturbseq/scripts/processing_script.sh b/wholebrainperturbseq/scripts/processing_script.sh
@@ -78,0 +78,22 @@
    MERGE_CMD="${MERGE_CMD} --data_path ${PROCESSED_DIR}/batch${i}/${SCREEN_NAME}_batch${i}_gex_filtered.h5ad"
done
-MERGE_CMD="${MERGE_CMD} --output_dir ${PROCESSED_DIR} --output_name ${SCREEN_NAME}_gex_filtered.h5ad" --write_zarr True --write_h5ad True
+MERGE_CMD="${MERGE_CMD} --output_dir ${PROCESSED_DIR} --output_name ${SCREEN_NAME}_gex_filtered.h5ad --write_h5ad"

echo "Executing: ${MERGE_CMD}"
eval ${MERGE_CMD}
echo "✓ Batches merged"
echo ""

# Step 4: Final QC on merged file
echo "Step 4: Final QC on merged file..."
python filter_for_deg.py \
    --data_path "${PROCESSED_DIR}/${SCREEN_NAME}_gex_filtered.h5ad" \
    --output_dir "${PROCESSED_DIR}/" \
    --output_name "${SCREEN_NAME}_gex_filtered.h5ad" \
    --min_cells_per_gene ${FINAL_MIN_CELLS_PER_GENE} \
    --min_cells_per_subclass ${FINAL_MIN_CELLS_PER_SUBCLASS} \
    --remove_unassigned \
    --unassigned_label "${UNASSIGNED_LABEL}" \
-    --write_zarr
+    --write_h5ad
echo "✓ Final QC completed"
```

This will make the zarr and h5ad flags work properly. Otherwise, if you try to write to zarr, it just won't work! (Speaking from experience here).

You will need to update the `BASE_DATA_PATH` in the script to wherever you want it to go to.

Now, it should work. I ran it with this command:

```bash
PYTHONPATH=/path/to/project/working/directory scripts/processing_script.sh 1
```

This will run one batch in processing. I tried all four batches, or even just two, but my computer crashes on the second batch. Fortunately, even with just 1 batch, I get >100,000 cells, which I'm happy with.

## Environment stuff

Make sure `py_config` is pointing to an environment that has anndata, otherwise reticulate will fail to load in the h5ad shards.

I think for computers with limited RAM, using zarr format in the download and analysis might be easier on your computer.

```{r starting}
py_config()
```

# Introduction

This dataset is a massive whole brain CRISPR screen. It follows the current trend from the Allen Institute, Arc, and others of pushing out massive datasets with perturbations to help build models and investigate interesting dynamics.

## setup pathways

```{r pathways_setup}
msig_mouse <- msigdbr(species = "Mus musculus")
colnames(msig_mouse)

hallmark_df <- msig_mouse %>%
  filter(gs_collection == "H")
pathways_h <- split(hallmark_df$gene_symbol, hallmark_df$gs_name)

# C2 curated pathways
c2_df <- msig_mouse %>%
  filter(gs_collection == "C2")
pathways_c2 <- split(c2_df$gene_symbol, c2_df$gs_name)

go_bp_df <- msig_mouse %>%
  filter(gs_subcollection == "GO:BP")
pathways_go_bp <- split(go_bp_df$gene_symbol, go_bp_df$gs_name)

go_cc_df <- msig_mouse %>%
  filter(gs_subcollection == "GO:CC")
pathways_go_cc <- split(go_cc_df$gene_symbol, go_cc_df$gs_name)

go_mf_df <- msig_mouse %>%
  filter(gs_subcollection == "GO:MF")
pathways_go_mf <- split(go_mf_df$gene_symbol, go_mf_df$gs_name)

pathways_list <- list(
  Hallmark = pathways_h,
  C2 = pathways_c2,
  GO_BP = pathways_go_bp,
  GO_CC = pathways_go_cc,
  GO_MF = pathways_go_mf
)
```

## load data

```{r data}
h5ad_path <- "/home/yoyomanzoor/Crucial/wholebrainperturbseq/processed/WB8588_screen_gex_filtered.h5ad"

sce <- readH5AD(h5ad_path)

sce
dim(sce)         # genes x cells
assayNames(sce)  # available assays
```

## metadata overview and quality control

```{r metadata}
# What metadata columns are available?
colnames(colData(sce))

colnames(colData(sce))  # cell metadata columns
rownames(rowData(sce))[1:5]
```

```{r stats}
cat("Cells:", ncol(sce), "\n")
cat("Genes:", nrow(sce), "\n")
cat("Unique mice:", length(unique(sce$mouse)), "\n")
cat("Unique targets:", length(unique(sce$gene_target)), "\n")
cat("Unique subclasses:", length(unique(sce$subclass_name)), "\n")
cat("Unique classes:", length(unique(sce$class_name)), "\n")
```

```{r metadata_summary}
sort(table(sce$class_name), decreasing = TRUE)[1:20]
sort(table(sce$subclass_name), decreasing = TRUE)[1:30]
sort(table(sce$cluster_name), decreasing = TRUE)[1:30]
sort(table(sce$gene_target), decreasing = TRUE)[1:30]
sort(table(sce$mouse), decreasing = TRUE)
sort(table(sce$sample_label), decreasing = TRUE)
sort(table(sce$region_level1), decreasing = TRUE)
sort(table(sce$region_level2), decreasing = TRUE)[1:30]
```

```{r qc_summary}
summary(sce$total_counts)
summary(sce$num_genes)
summary(sce$pct_counts_mt)
summary(sce$num_guides)
summary(sce$log_ambient_mse_norm)
summary(sce$predicted_subclass_probability)
```

```{r qc_plots}
qc_df <- as.data.frame(colData(sce))

p1 <- ggplot(qc_df, aes(x = total_counts)) +
  geom_histogram(bins = 100) +
  scale_x_log10() +
  ggtitle("Total counts per cell")

p2 <- ggplot(qc_df, aes(x = num_genes)) +
  geom_histogram(bins = 100) +
  scale_x_log10() +
  ggtitle("Detected genes per cell")

p3 <- ggplot(qc_df, aes(x = pct_counts_mt)) +
  geom_histogram(bins = 100) +
  ggtitle("Percent mitochondrial counts")

p4 <- ggplot(qc_df, aes(x = num_guides)) +
  geom_histogram(bins = 50) +
  ggtitle("Number of guides per cell")

(p1 | p2) / (p3 | p4)
```

```{r cell_type_tables}
class_counts <- as.data.frame(sort(table(sce$class_name), decreasing = TRUE))
colnames(class_counts) <- c("class_name", "n_cells")

subclass_counts <- as.data.frame(sort(table(sce$subclass_name), decreasing = TRUE))
colnames(subclass_counts) <- c("subclass_name", "n_cells")

cluster_counts <- as.data.frame(sort(table(sce$cluster_name), decreasing = TRUE))
colnames(cluster_counts) <- c("cluster_name", "n_cells")

head(class_counts, 20)
head(subclass_counts, 30)
head(cluster_counts, 30)
```

```{r perturbation_overview}
target_counts <- as.data.frame(sort(table(sce$gene_target), decreasing = TRUE))
colnames(target_counts) <- c("gene_target", "n_cells")

head(target_counts, 30)

# non-target vs others
table(sce$gene_target == "Non_target")
```

```{r cross_tab}
top_subclasses <- names(sort(table(sce$subclass_name), decreasing = TRUE))[1:15]
top_targets <- names(sort(table(sce$gene_target), decreasing = TRUE))[1:15]

xt <- table(
  subclass = factor(sce$subclass_name, levels = top_subclasses),
  target = factor(sce$gene_target, levels = top_targets)
)

xt
```

## UMAP and visualization

```{r umap}
assay(sce, "counts") <- assay(sce, "X")
set.seed(1)
idx <- sample(seq_len(ncol(sce)), min(20000, ncol(sce)))
sce_sub <- sce[, idx]
# normalize from counts assay
sce_sub <- logNormCounts(sce_sub)

# variable genes
dec <- modelGeneVar(sce_sub)
top_hvgs <- getTopHVGs(dec, n = 2000)

# PCA + UMAP
sce_sub <- runPCA(sce_sub, subset_row = top_hvgs, ncomponents = 30)
sce_sub <- runUMAP(sce_sub, dimred = "PCA")

# plotReducedDim(
#   sce_sub,
#   dimred = "UMAP",
#   colour_by = "subclass_name"
# ) + ggtitle("UMAP colored by subclass")

plotReducedDim(
  sce_sub,
  dimred = "UMAP",
  colour_by = "class_name"
) + ggtitle("UMAP colored by class")

top_targets <- names(sort(table(sce_sub$gene_target), decreasing = TRUE))[1:12]

sce_sub$gene_target_plot <- ifelse(
  sce_sub$gene_target %in% top_targets,
  sce_sub$gene_target,
  "other"
)

plotReducedDim(
  sce_sub,
  dimred = "UMAP",
  colour_by = "gene_target_plot"
) + ggtitle("UMAP colored by common perturbations")

plotReducedDim(
  sce_sub,
  dimred = "UMAP",
  colour_by = "mouse"
) + ggtitle("UMAP colored by mouse")

plotReducedDim(
  sce_sub,
  dimred = "UMAP",
  colour_by = "region_level1"
) + ggtitle("UMAP colored by region_level1")

plotReducedDim(
  sce_sub,
  dimred = "UMAP",
  colour_by = "region_level2"
) + ggtitle("UMAP colored by region_level2")
```

## Data investigation

I was interested in seeing if Pvalb, Sst, and Lamp5 interneurons had different cell types. I hopes they would be cortical and show ganglionic eminence differences, which they do!

```{r}
sce[, sce$subclass_name == "052 Pvalb Gaba"]$class_name %>% table() %>% sort(decreasing = TRUE)
sce[, sce$subclass_name == "053 Sst Gaba"]$class_name %>% table() %>% sort(decreasing = TRUE)
sce[, sce$subclass_name == "049 Lamp5 Gaba"]$class_name %>% table() %>% sort(decreasing = TRUE)
```

# Pvalb vs Lamp5

I first tried comparing these two subclasses using fgsea.

```{r subclass_analysis}
subclasses_keep <- c("052 Pvalb Gaba", "049 Lamp5 Gaba")

sce2 <- sce[, sce$subclass_name %in% subclasses_keep]
sce2$subclass_name <- droplevels(factor(sce2$subclass_name, levels = subclasses_keep))
table(sce2$subclass_name)

sce2$region_level1 %>% table() %>% sort(decreasing = TRUE)
sce2$class_name %>% table() %>% sort(decreasing = TRUE)

sce2_ctrl <- sce2[, sce2$gene_target == "Non_target"]
sce2_ctrl$subclass_name <- droplevels(sce2_ctrl$subclass_name)
table(sce2_ctrl$subclass_name)
```

```{r umap_subclasses}
sce_umap <- sce2_ctrl

sce_umap <- logNormCounts(sce_umap)

dec <- modelGeneVar(sce_umap)
top_hvgs <- getTopHVGs(dec, n = 2000)

sce_umap <- runPCA(sce_umap, subset_row = top_hvgs, ncomponents = 30)
sce_umap <- runUMAP(sce_umap, dimred = "PCA")

plotReducedDim(
  sce_umap,
  dimred = "UMAP",
  colour_by = "subclass_name"
) + ggtitle("Pvalb Gaba vs Lamp5 Gaba")

plotReducedDim(
  sce_umap,
  dimred = "UMAP",
  colour_by = "mouse"
) + ggtitle("UMAP colored by mouse")
```

```{r deg}
sce_de <- sce2_ctrl

pb_id <- interaction(sce_de$mouse, sce_de$subclass_name, drop = TRUE)

pb_counts <- sapply(levels(pb_id), function(g) {
  cols <- which(pb_id == g)
  Matrix::rowSums(assay(sce_de, "counts")[, cols, drop = FALSE])
})

pb_counts <- as.matrix(pb_counts)
colnames(pb_counts) <- levels(pb_id)

pb_meta <- data.frame(sample = colnames(pb_counts), stringsAsFactors = FALSE)
pb_meta$mouse <- sub("\\.(.*)$", "", pb_meta$sample)
pb_meta$subclass_name <- sub("^[^.]+\\.", "", pb_meta$sample)

pb_meta$subclass_name <- factor(
  pb_meta$subclass_name,
  levels = c("049 Lamp5 Gaba", "052 Pvalb Gaba")
)

pb_meta
```

```{r edgeR}
y <- DGEList(counts = pb_counts)
keep <- filterByExpr(y, group = pb_meta$subclass_name)
y <- y[keep, , keep.lib.sizes = FALSE]
y <- calcNormFactors(y)

design <- model.matrix(~ subclass_name, data = pb_meta)
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)
qlf <- glmQLFTest(fit, coef = 2)

de_tab <- topTags(qlf, n = Inf)$table
head(de_tab)
```

```{r ranks}
ranks <- de_tab$logFC * -log10(de_tab$PValue + 1e-300)
names(ranks) <- rownames(de_tab)
ranks <- sort(ranks, decreasing = TRUE)

head(ranks)
tail(ranks)
```

```{r inspect_de_tab}
de_tab %>%
  arrange(desc(logFC)) %>%
  head(20)

de_tab %>%
  arrange(logFC) %>%
  head(20)
```

```{r fgsea}
fg_h <- fgsea(
  pathways = pathways_h,
  stats = ranks,
  minSize = 15,
  maxSize = 500
)

fg_h <- fg_h[order(fg_h$padj), ]
head(fg_h[, c("pathway", "NES", "padj", "size")], 20)
```

```{r fgsea_c2}
fg_c2 <- fgsea(
  pathways = pathways_c2,
  stats = ranks,
  minSize = 15,
  maxSize = 500
)

fg_c2 <- fg_c2[order(fg_c2$padj), ]
head(fg_c2[, c("pathway", "NES", "padj", "size")], 20)
```

```{r fgsea_c5}
fg_go_bp <- fgsea(
  pathways = pathways_go_bp,
  stats = ranks,
  minSize = 15,
  maxSize = 500
)

fg_go_bp <- fg_go_bp[order(fg_go_bp$padj), ]
head(fg_go_bp[, c("pathway", "NES", "padj", "size")], 20)

top_fg_go_bp <- fg_go_bp %>%
  as.data.frame() %>%
  arrange(padj) %>%
  slice_head(n = 15)

ggplot(top_fg_go_bp, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = "Pvalb vs Lamp5: GO:BP",
    fill = "FDR < 0.05"
  ) +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))

fg_go_cc <- fgsea(
  pathways = pathways_go_cc,
  stats = ranks,
  minSize = 15,
  maxSize = 500
)

fg_go_cc <- fg_go_cc[order(fg_go_cc$padj), ]
head(fg_go_cc[, c("pathway", "NES", "padj", "size")], 20)

top_fg_go_cc <- fg_go_cc %>%
  as.data.frame() %>%
  arrange(padj) %>%
  slice_head(n = 15)

ggplot(top_fg_go_cc, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = "Pvalb vs Lamp5: GO:CC",
    fill = "FDR < 0.05"
  ) +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))

fg_go_mf <- fgsea(
  pathways = pathways_go_mf,
  stats = ranks,
  minSize = 15,
  maxSize = 500
)

fg_go_mf <- fg_go_mf[order(fg_go_mf$padj), ]
head(fg_go_mf[, c("pathway", "NES", "padj", "size")], 20)

top_fg_go_mf <- fg_go_mf %>%
  as.data.frame() %>%
  arrange(padj) %>%
  slice_head(n = 15)

ggplot(top_fg_go_mf, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = "Pvalb vs Lamp5: GO:MF",
    fill = "FDR < 0.05"
  ) +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))
```

```{r plot_fgsea}
#| out-width: "60%"
top_fg <- fg_h %>%
  as.data.frame() %>%
  arrange(padj) %>%
  slice_head(n = 15)

ggplot(top_fg, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = "Pvalb vs Lamp5: Hallmark"
  ) +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))

top_fg_c2 <- fg_c2 %>%
  as.data.frame() %>%
  arrange(padj) %>%
  slice_head(n = 15)

ggplot(top_fg_c2, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = "Pvalb vs Lamp5: C2 curated pathways"
  ) +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))
```

# Pvalb vs Sst

Then I interrogated Pvalb and Sst in the same way.

```{r subclass_pvalb_sst}
subclasses_keep <- c("052 Pvalb Gaba", "053 Sst Gaba")

sce2 <- sce[, sce$subclass_name %in% subclasses_keep]
sce2$subclass_name <- droplevels(factor(sce2$subclass_name, levels = subclasses_keep))
table(sce2$subclass_name)

sce2$region_level1 %>% table() %>% sort(decreasing = TRUE)
```

These cells are entirely cortical, as expected. 

```{r}
#| out-width: "60%"
sce2_ctrl <- sce2[, sce2$gene_target == "Non_target"]
sce2_ctrl$subclass_name <- droplevels(sce2_ctrl$subclass_name)
table(sce2_ctrl$subclass_name)

sce_umap <- sce2_ctrl

sce_umap <- logNormCounts(sce_umap)

dec <- modelGeneVar(sce_umap)
top_hvgs <- getTopHVGs(dec, n = 2000)

sce_umap <- runPCA(sce_umap, subset_row = top_hvgs, ncomponents = 30)
sce_umap <- runUMAP(sce_umap, dimred = "PCA")

plotReducedDim(
  sce_umap,
  dimred = "UMAP",
  colour_by = "subclass_name"
) + ggtitle("Pvalb Gaba vs Sst Gaba")

plotReducedDim(
  sce_umap,
  dimred = "UMAP",
  colour_by = "mouse"
) + ggtitle("UMAP colored by mouse")

sce_de <- sce2_ctrl

pb_id <- interaction(sce_de$mouse, sce_de$subclass_name, drop = TRUE)

pb_counts <- sapply(levels(pb_id), function(g) {
  cols <- which(pb_id == g)
  Matrix::rowSums(assay(sce_de, "counts")[, cols, drop = FALSE])
})

pb_counts <- as.matrix(pb_counts)
colnames(pb_counts) <- levels(pb_id)

pb_meta <- data.frame(sample = colnames(pb_counts), stringsAsFactors = FALSE)
pb_meta$mouse <- sub("\\.(.*)$", "", pb_meta$sample)
pb_meta$subclass_name <- sub("^[^.]+\\.", "", pb_meta$sample)

pb_meta$subclass_name <- factor(
  pb_meta$subclass_name,
  levels = c("052 Pvalb Gaba", "053 Sst Gaba")
)

pb_meta

y <- DGEList(counts = pb_counts)
keep <- filterByExpr(y, group = pb_meta$subclass_name)
y <- y[keep, , keep.lib.sizes = FALSE]
y <- calcNormFactors(y)

design <- model.matrix(~ subclass_name, data = pb_meta)
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)
qlf <- glmQLFTest(fit, coef = 2)

de_tab <- topTags(qlf, n = Inf)$table
head(de_tab)

ranks <- de_tab$logFC * -log10(de_tab$PValue + 1e-300)
names(ranks) <- rownames(de_tab)
ranks <- sort(ranks, decreasing = TRUE)

head(ranks)
tail(ranks)

fg_h <- fgsea(
  pathways = pathways_h,
  stats = ranks,
  minSize = 15,
  maxSize = 500
)

fg_h <- fg_h[order(fg_h$padj), ]
head(fg_h[, c("pathway", "NES", "padj", "size")], 20)

top_fg <- fg_h %>%
  as.data.frame() %>%
  arrange(padj) %>%
  slice_head(n = 15)

ggplot(top_fg, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = "Pvalb vs Sst: Hallmark"
  ) +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))

fg_c2 <- fgsea(
  pathways = pathways_c2,
  stats = ranks,
  minSize = 15,
  maxSize = 500
)

fg_c2 <- fg_c2[order(fg_c2$padj), ]
head(fg_c2[, c("pathway", "NES", "padj", "size")], 20)

top_fg_c2 <- fg_c2 %>%
  as.data.frame() %>%
  arrange(padj) %>%
  slice_head(n = 15)

ggplot(top_fg_c2, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = "Pvalb vs Sst: C2 curated pathways"
  ) +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))

fg_go_bp <- fgsea(
  pathways = pathways_go_bp,
  stats = ranks,
  minSize = 15,
  maxSize = 500
)

fg_go_bp <- fg_go_bp[order(fg_go_bp$padj), ]
head(fg_go_bp[, c("pathway", "NES", "padj", "size")], 20)

top_fg_go_bp <- fg_go_bp %>%
  as.data.frame() %>%
  arrange(padj) %>%
  slice_head(n = 15)

ggplot(top_fg_go_bp, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = "Pvalb vs Sst: GO:BP"
  ) +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))

fg_go_cc <- fgsea(
  pathways = pathways_go_cc,
  stats = ranks,
  minSize = 15,
  maxSize = 500
)

fg_go_cc <- fg_go_cc[order(fg_go_cc$padj), ]
head(fg_go_cc[, c("pathway", "NES", "padj", "size")], 20)

top_fg_go_cc <- fg_go_cc %>%
  as.data.frame() %>%
  arrange(padj) %>%
  slice_head(n = 5)

ggplot(top_fg_go_cc, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = "Pvalb vs Sst: GO:CC"
  ) +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))

fg_go_mf <- fgsea(
  pathways = pathways_go_mf,
  stats = ranks,
  minSize = 15,
  maxSize = 500
)

fg_go_mf <- fg_go_mf[order(fg_go_mf$padj), ]
head(fg_go_mf[, c("pathway", "NES", "padj", "size")], 20)

top_fg_go_mf <- fg_go_mf %>%
  as.data.frame() %>%
  arrange(padj) %>%
  slice_head(n = 15)

ggplot(top_fg_go_mf, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = "Pvalb vs Sst: GO:MF"
  ) +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))

top_fg_go_mf %>% arrange(NES)
```

# Perturbation investigation

This function was generated with the help of AI, based on the analysis so far.

```{r}
run_multi_fgsea <- function(
  sce,
  pathways_list,
  group_col,
  group1,
  group2,
  sample_col = "mouse",
  gene_target = NULL,
  region_level1 = NULL,
  region_level2 = NULL,
  class_name = NULL,
  subclass_name = NULL,
  minSize = 15,
  maxSize = 500,
  top_n = 15,
  n_hvgs = 2000,
  n_pcs = 30,
  rank_fun = NULL
) {
  # Required packages:
  # library(SingleCellExperiment)
  # library(scuttle)
  # library(scran)
  # library(scater)
  # library(edgeR)
  # library(fgsea)
  # library(dplyr)
  # library(ggplot2)
  # library(Matrix)

  if (is.null(rank_fun)) {
    rank_fun <- function(de_tab) {
      r <- de_tab$logFC * -log10(de_tab$PValue + 1e-300)
      names(r) <- rownames(de_tab)
      sort(r, decreasing = TRUE)
    }
  }

  #----------------------------
  # 1) Filter cells
  #----------------------------
  sce_sub <- sce

  if (!is.null(gene_target)) {
    sce_sub <- sce_sub[, sce_sub$gene_target %in% gene_target]
  }
  if (!is.null(region_level1)) {
    sce_sub <- sce_sub[, sce_sub$region_level1 %in% region_level1]
  }
  if (!is.null(region_level2)) {
    sce_sub <- sce_sub[, sce_sub$region_level2 %in% region_level2]
  }
  if (!is.null(class_name)) {
    sce_sub <- sce_sub[, sce_sub$class_name %in% class_name]
  }
  if (!is.null(subclass_name)) {
    sce_sub <- sce_sub[, sce_sub$subclass_name %in% subclass_name]
  }

  if (ncol(sce_sub) == 0) {
    stop("No cells remain after filtering.")
  }

  if (!group_col %in% colnames(colData(sce_sub))) {
    stop("group_col not found in colData(sce).")
  }
  if (!sample_col %in% colnames(colData(sce_sub))) {
    stop("sample_col not found in colData(sce).")
  }

  sce_sub <- sce_sub[, sce_sub[[group_col]] %in% c(group1, group2)]

  if (ncol(sce_sub) == 0) {
    stop("No cells remain after restricting to group1/group2.")
  }

  sce_sub[[group_col]] <- droplevels(factor(sce_sub[[group_col]], levels = c(group1, group2)))

  present_groups <- levels(droplevels(sce_sub[[group_col]]))
  if (!all(c(group1, group2) %in% present_groups)) {
    stop("Both group1 and group2 must be present after filtering.")
  }

  #----------------------------
  # 2) UMAP workflow
  #----------------------------
  sce_umap <- sce_sub
  sce_umap <- scuttle::logNormCounts(sce_umap)

  dec <- scran::modelGeneVar(sce_umap)
  top_hvgs <- scran::getTopHVGs(dec, n = n_hvgs)

  sce_umap <- scater::runPCA(sce_umap, subset_row = top_hvgs, ncomponents = n_pcs)
  sce_umap <- scater::runUMAP(sce_umap, dimred = "PCA")

  umap_coords <- as.data.frame(SingleCellExperiment::reducedDim(sce_umap, "UMAP"))
  colnames(umap_coords) <- c("UMAP1", "UMAP2")

  umap_df <- cbind(
    umap_coords,
    as.data.frame(colData(sce_umap))
  ) %>%
    dplyr::mutate(
      plot_group = .data[[group_col]]
    )

  #----------------------------
  # 3) Pseudobulk
  #----------------------------
  pb_id <- interaction(
    sce_sub[[sample_col]],
    sce_sub[[group_col]],
    drop = TRUE
  )

  pb_counts <- sapply(levels(pb_id), function(g) {
    cols <- which(pb_id == g)
    Matrix::rowSums(SummarizedExperiment::assay(sce_sub, "counts")[, cols, drop = FALSE])
  })

  pb_counts <- as.matrix(pb_counts)
  colnames(pb_counts) <- levels(pb_id)

  pb_meta <- data.frame(sample = colnames(pb_counts), stringsAsFactors = FALSE)
  pb_meta[[sample_col]] <- sub("\\.(.*)$", "", pb_meta$sample)
  pb_meta[[group_col]] <- sub("^[^.]+\\.", "", pb_meta$sample)
  pb_meta[[group_col]] <- factor(pb_meta[[group_col]], levels = c(group1, group2))

  #----------------------------
  # 4) Differential expression
  #----------------------------
  y <- edgeR::DGEList(counts = pb_counts)
  keep <- edgeR::filterByExpr(y, group = pb_meta[[group_col]])
  y <- y[keep, , keep.lib.sizes = FALSE]
  y <- edgeR::calcNormFactors(y)

  design_df <- pb_meta
  design <- stats::model.matrix(
    stats::as.formula(paste0("~ ", group_col)),
    data = design_df
  )

  y <- edgeR::estimateDisp(y, design)
  fit <- edgeR::glmQLFit(y, design)
  qlf <- edgeR::glmQLFTest(fit, coef = 2)

  de_tab <- edgeR::topTags(qlf, n = Inf)$table
  ranks <- rank_fun(de_tab)

  #----------------------------
  # 5) FGSEA across databases
  #----------------------------
  fgsea_list <- lapply(names(pathways_list), function(db_name) {
    fg <- fgsea::fgsea(
      pathways = pathways_list[[db_name]],
      stats = ranks,
      minSize = minSize,
      maxSize = maxSize
    )

    fg <- fg[order(fg$padj), ]
    fg_df <- as.data.frame(fg) %>%
      dplyr::mutate(
        database = db_name,
        contrast = paste0(group2, " vs ", group1)
      )

    top_df <- fg_df %>%
      dplyr::arrange(padj) %>%
      dplyr::slice_head(n = top_n) %>%
      dplyr::mutate(
        pathway = factor(pathway, levels = pathway[order(NES)])
      )

    list(
      full = fg_df,
      top = top_df
    )
  })

  names(fgsea_list) <- names(pathways_list)

  fgsea_all <- dplyr::bind_rows(lapply(fgsea_list, `[[`, "full")) %>%
    dplyr::arrange(database, padj)

  fgsea_top <- dplyr::bind_rows(lapply(fgsea_list, `[[`, "top")) %>%
    dplyr::arrange(database, padj)

  #----------------------------
  # 6) Return
  #----------------------------
  list(
    sce_filtered = sce_sub,
    sce_umap = sce_umap,
    umap_df = umap_df,
    pb_meta = pb_meta,
    de_table = de_tab,
    ranks = ranks,
    fgsea = fgsea_list,
    fgsea_all = fgsea_all,
    fgsea_top = fgsea_top,
    contrast = list(
      group_col = group_col,
      group1 = group1,
      group2 = group2
    )
  )
}

get_leading_edge <- function(res, database, target_pathway) {
  df <- res$fgsea[[database]]$full
  
  genes <- df %>%
    dplyr::filter(pathway == target_pathway) %>%
    dplyr::pull(leadingEdge) %>%
    .[[1]]
  
  out <- res$de_table[genes, , drop = FALSE]
  out$gene <- rownames(out)
  
  out %>%
    dplyr::arrange(desc(logFC))
}
```

```{r}
res <- run_multi_fgsea(
  sce = sce,
  pathways_list = pathways_list,
  group_col = "gene_target",
  group1 = "Non_target",
  group2 = "Mif",
  region_level1 = "Cortex"
)
```

```{r}
ggplot(res$umap_df, aes(x = UMAP1, y = UMAP2, color = plot_group)) +
  geom_point(alpha = 0.7) +
  labs(
    color = res$contrast$group_col,
    title = paste0("UMAP colored by ", res$contrast$group_col)
  ) +
  scale_color_manual(values = c(`Non_target` = "#CCCCCC", `Mif` = "Black"))
```

```{r}
res$fgsea_top %>%
  dplyr::filter(database == "GO_BP") %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = paste0("FGSEA: ", res$contrast$group2, " vs ", res$contrast$group1, ": GO_BP")
  ) +
  theme_bw() +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))
```

```{r}
ggplot(res$fgsea_top, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ database, scales = "free_y") +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = paste0("FGSEA: ", res$contrast$group2, " vs ", res$contrast$group1)
  ) +
  theme_bw() +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))
```

```{r}
top_fg_h <- res$fgsea$Hallmark$top

ggplot(top_fg_h, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = paste0("FGSEA: ", res$contrast$group2, " vs ", res$contrast$group1, ": Hallmark")
  ) +
  theme_bw() +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))
```

```{r}
res_pou3f3 <- run_multi_fgsea(
  sce = sce,
  pathways_list = pathways_list,
  group_col = "gene_target",
  group1 = "Non_target",
  group2 = "Pou3f3",
  region_level1 = "Cortex"
)

ggplot(res_pou3f3$umap_df, aes(x = UMAP1, y = UMAP2, color = plot_group)) +
  geom_point(alpha = 0.7) +
  labs(
    color = res_pou3f3$contrast$group_col,
    title = paste0("UMAP colored by ", res_pou3f3$contrast$group_col)
  ) +
  scale_color_manual(values = c(`Non_target` = "#CCCCCC", `Pou3f3` = "Black"))

res_pou3f3$fgsea_top %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = paste0("FGSEA: ", res_pou3f3$contrast$group2, " vs ", res_pou3f3$contrast$group1)
  ) +
  theme_bw() +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))


top_fg_h <- res_pou3f3$fgsea$Hallmark$top

ggplot(top_fg_h, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = paste0("FGSEA: ", res_pou3f3$contrast$group2, " vs ", res_pou3f3$contrast$group1, ": Hallmark")
  ) +
  theme_bw() +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))
```

```{r}
res_zbtb7a <- run_multi_fgsea(
  sce = sce,
  pathways_list = pathways_list,
  group_col = "gene_target",
  group1 = "Non_target",
  group2 = "Zbtb7a",
  region_level1 = "Cortex"
)

ggplot(res_zbtb7a$umap_df, aes(x = UMAP1, y = UMAP2, color = plot_group)) +
  geom_point(alpha = 0.7) +
  labs(
    color = res_zbtb7a$contrast$group_col,
    title = paste0("UMAP colored by ", res_zbtb7a$contrast$group_col)
  ) +
  scale_color_manual(values = c(`Non_target` = "#CCCCCC", `Zbtb7a` = "Black"))

res_zbtb7a$fgsea_top %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = paste0("FGSEA: ", res_zbtb7a$contrast$group2, " vs ", res_zbtb7a$contrast$group1)
  ) +
  theme_bw() +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))

top_fg_h <- res_zbtb7a$fgsea$Hallmark$top

ggplot(top_fg_h, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = paste0("FGSEA: ", res_zbtb7a$contrast$group2, " vs ", res_zbtb7a$contrast$group1, ": Hallmark")
  ) +
  theme_bw() +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))
```

```{r}
res_zbtb7a_vs_adrb3 <- run_multi_fgsea(
  sce = sce,
  pathways_list = pathways_list,
  group_col = "gene_target",
  group1 = "Adrb3",
  group2 = "Zbtb7a",
  region_level1 = "Cortex"
)

ggplot(res_zbtb7a_vs_adrb3$umap_df, aes(x = UMAP1, y = UMAP2, color = plot_group)) +
  geom_point(alpha = 0.7) +
  labs(
    color = res_zbtb7a_vs_adrb3$contrast$group_col,
    title = paste0("UMAP colored by ", res_zbtb7a_vs_adrb3$contrast$group_col)
  )

res_zbtb7a_vs_adrb3$fgsea_top %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = paste0("FGSEA: ", res_zbtb7a_vs_adrb3$contrast$group2, " vs ", res_zbtb7a_vs_adrb3$contrast$group1)
  ) +
  theme_bw() +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))


top_fg_h <- res_zbtb7a_vs_adrb3$fgsea$Hallmark$top

ggplot(top_fg_h, aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = paste0("FGSEA: ", res_zbtb7a_vs_adrb3$contrast$group2, " vs ", res_zbtb7a_vs_adrb3$contrast$group1, ": Hallmark")
  ) +
  theme_bw() +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))
```

```{r}
res_pou3f3_vs_mif <- run_multi_fgsea(
  sce = sce,
  pathways_list = pathways_list,
  group_col = "gene_target",
  group1 = "Mif",
  group2 = "Pou3f3",
  region_level1 = "Cortex"
)

ggplot(res_pou3f3_vs_mif$umap_df, aes(x = UMAP1, y = UMAP2, color = plot_group)) +
  geom_point(alpha = 0.7) +
  labs(
    color = res_pou3f3_vs_mif$contrast$group_col,
    title = paste0("UMAP colored by ", res_pou3f3_vs_mif$contrast$group_col)
  )

res_pou3f3_vs_mif$fgsea_top %>%
  dplyr::filter(database == "GO_BP") %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = paste0("FGSEA: ", res_pou3f3_vs_mif$contrast$group2, " vs ", res_pou3f3_vs_mif$contrast$group1, ": GO_BP")
  ) +
  theme_bw() +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))

res_pou3f3_vs_mif$fgsea_top %>%
  dplyr::filter(database == "C2") %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = paste0("FGSEA: ", res_pou3f3_vs_mif$contrast$group2, " vs ", res_pou3f3_vs_mif$contrast$group1, ": C2")
  ) +
  theme_bw() +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))

get_leading_edge(res_pou3f3_vs_mif, "C2", "WP_OXIDATIVE_PHOSPHORYLATION")

le <- get_leading_edge(res_pou3f3_vs_mif, "C2", "WP_SPINAL_CORD_INJURY")

genes <- le %>%
  dplyr::distinct(gene) |>
  dplyr::filter(!is.na(gene)) |>
  dplyr::filter(gene %in% rownames(res_pou3f3_vs_mif$sce_umap)) |>
  dplyr::slice_head(n = 20) |>
  dplyr::pull(gene)

scater::plotExpression(
  res_pou3f3_vs_mif$sce_umap,
  features = genes,
  x = res_pou3f3_vs_mif$contrast$group_col
)
```

```{r}
res_sox11_vs_gata6 <- run_multi_fgsea(
  sce = sce,
  pathways_list = pathways_list,
  group_col = "gene_target",
  group1 = "Gata6",
  group2 = "Sox11",
)

p1 <-ggplot(res_sox11_vs_gata6$umap_df, aes(x = UMAP1, y = UMAP2, color = plot_group)) +
  geom_point(alpha = 0.7) +
  labs(
    color = res_sox11_vs_gata6$contrast$group_col,
    title = paste0("UMAP colored by ", res_sox11_vs_gata6$contrast$group_col)
  ) +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = "none")

p2 <- ggplot(res_sox11_vs_gata6$umap_df, aes(x = UMAP1, y = UMAP2, color = class_name)) +
  geom_point(alpha = 0.7) +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = "none") +
  ggtitle("UMAP colored by class")

p3 <- ggplot(res_sox11_vs_gata6$umap_df, aes(x = UMAP1, y = UMAP2, color = region_level1)) +
  geom_point(alpha = 0.7) +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = "none") +
  ggtitle("UMAP colored by region")

p <- (p1 | p2 | p3)
p

res_sox11_vs_gata6$fgsea_top %>%
  dplyr::filter(database == "GO_BP") %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = paste0("FGSEA: ", res_sox11_vs_gata6$contrast$group2, " vs ", res_sox11_vs_gata6$contrast$group1)
  ) +
  theme_bw() +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))
```

```{r}
res_pou3f3_vs_gata6 <- run_multi_fgsea(
  sce = sce,
  pathways_list = pathways_list,
  group_col = "gene_target",
  group1 = "Gata6",
  group2 = "Pou3f3",
)

ggplot(res_pou3f3_vs_gata6$umap_df, aes(x = UMAP1, y = UMAP2, color = plot_group)) +
  geom_point(alpha = 0.7) +
  labs(
    color = res_pou3f3_vs_gata6$contrast$group_col
  ) +
  ggtitle("UMAP colored by gene target")

res_pou3f3_vs_gata6$fgsea_top %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = paste0("FGSEA: ", res_pou3f3_vs_gata6$contrast$group2, " vs ", res_pou3f3_vs_gata6$contrast$group1)
  ) +
  theme_bw() +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))
```

```{r}
res_gnas_vs_zbtb7a <- run_multi_fgsea(
  sce = sce,
  pathways_list = pathways_list,
  group_col = "gene_target",
  group1 = "Zbtb7a",
  group2 = "Gnas",
)

ggplot(res_gnas_vs_zbtb7a$umap_df, aes(x = UMAP1, y = UMAP2, color = plot_group)) +
  geom_point(alpha = 0.7) +
  labs(
    color = res_gnas_vs_zbtb7a$contrast$group_col,
    title = paste0("UMAP colored by ", res_gnas_vs_zbtb7a$contrast$group_col)
  )

p <- res_gnas_vs_zbtb7a$fgsea_top %>%
  dplyr::filter(database == "GO_MF") %>%
  arrange(padj) %>%
  slice_head(n = 5) %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = padj < 0.05)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = paste0("FGSEA: ", res_gnas_vs_zbtb7a$contrast$group2, " vs ", res_gnas_vs_zbtb7a$contrast$group1)
  ) +
  theme_bw() +
  scale_fill_manual(values = c(`TRUE` = "#FFCB05", `FALSE` = "#CCCCCC"))
```

```{r}
le <- get_leading_edge(res_gnas_vs_zbtb7a, "GO_MF", "GOMF_TRANSCRIPTION_FACTOR_BINDING")

genes <- le %>%
  dplyr::distinct(gene) |>
  dplyr::filter(!is.na(gene)) |>
  dplyr::filter(gene %in% rownames(res_gnas_vs_zbtb7a$sce_umap)) |>
  dplyr::slice_head(n = 20) |>
  dplyr::pull(gene)

scater::plotExpression(
  res_gnas_vs_zbtb7a$sce_umap,
  features = genes,
  x = res_gnas_vs_zbtb7a$contrast$group_col
)
```

```{r}
le <- get_leading_edge(
  res_gnas_vs_zbtb7a,
  "GO_MF",
  "GOMF_TRANSCRIPTION_FACTOR_BINDING"
)

genes <- le %>%
  distinct(gene, .keep_all = TRUE) %>%
  filter(!is.na(gene)) %>%
  filter(gene %in% rownames(res_gnas_vs_zbtb7a$sce_umap)) %>%
  arrange(desc(abs(logFC))) %>%
  slice_head(n = 20) %>%
  pull(gene)

expr_mat <- SummarizedExperiment::assay(
  res_gnas_vs_zbtb7a$sce_umap,
  "logcounts"
)[genes, , drop = FALSE]

group_vec <- colData(res_gnas_vs_zbtb7a$sce_umap)[[res_gnas_vs_zbtb7a$contrast$group_col]]
names(group_vec) <- colnames(res_gnas_vs_zbtb7a$sce_umap)

heat_df <- as.data.frame(as.matrix(expr_mat)) %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "cell", values_to = "expr") %>%
  mutate(group = group_vec[cell]) %>%
  group_by(gene, group) %>%
  summarize(expr = mean(expr), .groups = "drop")

heat_mat <- heat_df %>%
  pivot_wider(names_from = group, values_from = expr) %>%
  as.data.frame()

rownames(heat_mat) <- heat_mat$gene
heat_mat$gene <- NULL
heat_mat <- as.matrix(heat_mat)

heat_mat <- heat_mat[, c(
  res_gnas_vs_zbtb7a$contrast$group1,
  res_gnas_vs_zbtb7a$contrast$group2
), drop = FALSE]

p <- pheatmap(
  t(heat_mat),
  scale = "none",
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  border_color = NA,
  fontsize_row = 10,
  fontsize_col = 11,
  main = "Zbtb7a vs Gnas: GOMF_TRANSCRIPTION_FACTOR_BINDING", # make main smaller
  main_font_size = 7
)
```

# Old code

```r
sce_pou3f3 <- sce[, sce$gene_target %in% c("Pou3f3", "Non_target")]
# table(sce_pou3f3$gene_target)

sce_pou3f3_pvalb <- sce[, sce$gene_target %in% c("Pou3f3", "Non_target") &
                        sce$subclass_name == "052 Pvalb Gaba"]

table(sce_pou3f3_pvalb$gene_target) %>% sort(decreasing = TRUE) %>% head(30)
```

```r umap_pou3f3_pvalb
# UMAP
sce_pou3f3_pvalb <- logNormCounts(sce_pou3f3_pvalb)
sce_pou3f3_pvalb <- runPCA(sce_pou3f3_pvalb, ncomponents = 30)
sce_pou3f3_pvalb <- runUMAP(sce_pou3f3_pvalb, dimred = "PCA")

plotReducedDim(
  sce_pou3f3_pvalb,
  dimred = "UMAP",
  colour_by = "gene_target"
) +
  scale_color_manual(values = c("Non_target" = "#d9d9d9", "Pou3f3" = "#1f1f1f")) +
  ggtitle("UMAP colored by gene target")
```

```r
check_reps <- function(sce, celltype) {
  as.data.frame(colData(sce)) %>%
    filter(subclass_name == celltype, passes_qc) %>%
    mutate(replicate_id = interaction(mouse, sample_label, drop = TRUE)) %>%
    summarise(
      n_cells = n(),
      n_mice = n_distinct(mouse),
      n_samples = n_distinct(sample_label),
      n_replicates = n_distinct(replicate_id)
    )
}
check_reps(sce, "052 Pvalb Gaba")
check_reps(sce, "049 Lamp5 Gaba")

check_nt <- function(sce, celltype) {
  as.data.frame(colData(sce)) %>%
    filter(subclass_name == celltype, passes_qc) %>%
    dplyr::count(gene_target, sort = TRUE)
}

check_nt(sce, "052 Pvalb Gaba") %>% head(20)
check_nt(sce, "049 Lamp5 Gaba") %>% head(20)

raw_pert_summary <- function(sce, celltype) {
  as.data.frame(colData(sce)) %>%
    filter(subclass_name == celltype, passes_qc) %>%
    mutate(replicate_id = interaction(mouse, sample_label, drop = TRUE)) %>%
    dplyr::count(gene_target, replicate_id, name = "n_cells") %>%
    group_by(gene_target) %>%
    summarise(
      n_replicates = n(),
      total_cells = sum(n_cells),
      min_cells = min(n_cells),
      max_cells = max(n_cells),
      .groups = "drop"
    ) %>%
    arrange(desc(n_replicates), desc(total_cells))
}

raw_pert_summary(sce, "052 Pvalb Gaba") %>% head(30)
raw_pert_summary(sce, "049 Lamp5 Gaba") %>% head(30)
```

This was code to help find perturbations that are located in a specific cell type through some form of annotation.

```r
scan_group_perturb_support <- function(
  sce,
  group_col,
  perturb_col = "gene_target",
  replicate_col = c("mouse", "mouse_sample"),
  qc_col = "passes_qc",
  non_target_label = "Non_target",
  min_cells_per_rep = 10,
  min_matched_reps = 3
) {
  replicate_col <- match.arg(replicate_col)

  meta <- as.data.frame(colData(sce))

  if (replicate_col == "mouse") {
    meta$replicate_id <- meta$mouse
  } else {
    meta$replicate_id <- interaction(meta$mouse, meta$sample_label, drop = TRUE)
  }

  meta <- meta %>%
    dplyr::filter(.data[[qc_col]]) %>%
    dplyr::filter(!is.na(.data[[group_col]]), !is.na(.data[[perturb_col]]))

  counts_df <- meta %>%
    dplyr::count(.data[[group_col]], .data[[perturb_col]], replicate_id, name = "n_cells") %>%
    dplyr::rename(group_value = 1, gene_target = 2)

  nt_df <- counts_df %>%
    dplyr::filter(gene_target == non_target_label) %>%
    dplyr::select(group_value, replicate_id, nt_cells = n_cells)

  counts_df %>%
    dplyr::filter(gene_target != non_target_label) %>%
    dplyr::inner_join(nt_df, by = c("group_value", "replicate_id")) %>%
    dplyr::mutate(
      ok_pair = n_cells >= min_cells_per_rep & nt_cells >= min_cells_per_rep
    ) %>%
    dplyr::group_by(group_value, gene_target) %>%
    dplyr::summarise(
      n_matched_replicates = sum(ok_pair),
      total_pert_cells = sum(n_cells),
      total_nt_cells = sum(nt_cells),
      min_pert_cells = min(n_cells),
      max_pert_cells = max(n_cells),
      min_nt_cells = min(nt_cells),
      max_nt_cells = max(nt_cells),
      .groups = "drop"
    ) %>%
    dplyr::filter(n_matched_replicates >= min_matched_reps) %>%
    dplyr::arrange(dplyr::desc(n_matched_replicates), dplyr::desc(total_pert_cells))
}

cand_subclass <- scan_group_perturb_support(
  sce,
  group_col = "subclass_name",
  replicate_col = "mouse",
  min_cells_per_rep = 10,
  min_matched_reps = 3
)

cand_class <- scan_group_perturb_support(
  sce,
  group_col = "class_name",
  replicate_col = "mouse",
  min_cells_per_rep = 10,
  min_matched_reps = 3
)

cand_region1 <- scan_group_perturb_support(
  sce,
  group_col = "region_level1",
  replicate_col = "mouse",
  min_cells_per_rep = 10,
  min_matched_reps = 3
)

cand_region2 <- scan_group_perturb_support(
  sce,
  group_col = "region_level2",
  replicate_col = "mouse",
  min_cells_per_rep = 10,
  min_matched_reps = 3
)

head(cand_subclass, 30)
head(cand_class, 30)
head(cand_region1, 30)
head(cand_region2, 30)
```


```r
library(Matrix)
perturbation <- "Zbtb7a"
region <- "Cortex"
sce_fgsea <- sce[, 
  sce$region_level1 == region &
  sce$gene_target %in% c("Non_target", perturbation) &
  sce$passes_qc
]

table(sce_fgsea$gene_target) %>% head(20)
table(sce_fgsea$mouse)

pb_id <- interaction(sce_fgsea$mouse, sce_fgsea$gene_target, drop = TRUE)

pb_counts <- sapply(levels(pb_id), function(g) {
  cols <- which(pb_id == g)
  Matrix::rowSums(assay(sce_fgsea, "counts")[, cols, drop = FALSE])
})

pb_counts <- as.matrix(pb_counts)
colnames(pb_counts) <- levels(pb_id)

pb_meta <- data.frame(sample = colnames(pb_counts))
pb_meta$mouse <- sub("\\.(.*)$", "", pb_meta$sample)
pb_meta$gene_target <- sub("^[^.]+\\.", "", pb_meta$sample)

pb_meta$gene_target <- factor(pb_meta$gene_target,
                             levels = c("Non_target", perturbation))

pb_meta

y <- DGEList(counts = pb_counts)

keep <- filterByExpr(y, group = pb_meta$gene_target)
y <- y[keep, , keep.lib.sizes = FALSE]

y <- calcNormFactors(y)

design <- model.matrix(~ gene_target, data = pb_meta)

y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)
qlf <- glmQLFTest(fit, coef = 2)

de_tab <- topTags(qlf, n = Inf)$table
head(de_tab)
ranks <- de_tab$logFC * -log10(de_tab$PValue + 1e-300)
names(ranks) <- rownames(de_tab)
ranks <- sort(ranks, decreasing = TRUE)
names(ranks) <- toupper(names(ranks))

msig_mouse <- msigdbr(species = "Mus musculus")
colnames(msig_mouse)

hallmark_df <- msig_mouse %>%
  filter(gs_collection == "H")

pathways <- split(hallmark_df$gene_symbol, hallmark_df$gs_name)
pathways <- lapply(pathways, toupper)

fg <- fgsea(
  pathways = pathways,
  stats = ranks,
  minSize = 15,
  maxSize = 500
)

fg <- fg[order(fg$padj), ]

head(fg[, c("pathway", "NES", "padj")], 20)

top_fg <- fg %>%
  as.data.frame() %>%
  arrange(padj) %>%
  slice_head(n = 15)

ggplot(top_fg, aes(x = reorder(pathway, NES), y = NES)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Normalized enrichment score",
    title = "FGSEA: Zbtb7a vs Non_target: Hallmark"
  )
```

